0/1 Knapsack

### 2015-03-13 Programming Project C/01knapsack.tex
should be complete

%% BEGIN ABSTRACT
\begin{abstract}
%%TODO: write this

@@

%% BEGIN MOTIVATION
\section{Motivation}
A 0/1 Knapsack problem is most often described as how does a theif steal the most valuable items with the amount of space he has available to carry them. However, the applications of a 0/1 Knapsack far beyond just this. When solved, a 0/1 Knapsack problem will render the optimal solution, whether it be how one should construct their schedule so they can get the most done in one day, or how to pack a truck such that the fewest amount of trips are taken.

@@

%% BEGIN BACKGROUND
\section{Background}
Unfortunately, solving a 0/1 Knapsack problem often times proves difficult because in order to find the best solution, all possible solutions must be attempted, a method known as brute force. It is not pratical to brute force a problem of this kind because it would take $ \theta{n!} $ to solve, which is unacceptable for large sets of numbers. However, there are other methods beside brute force that allow for the correct or 'good-enought' solution to be found. One of these methods is known as dynamic programming, which trades off speed for space. With dynamic programming the correct solution can be found in $ \theta{n*m} $ time, where n is the amount of items and m is the size of the knapsack. Another method for solving this type of problem is called a greedy algorithm, which works by making the best decision on which item to take locally. A greedy solution to the 0/1 Knapsack problem traditionally runs in $ \theta{n} $ time. Unfortunately a greedy solution is not always correct and therefore highly unreliable.

@@

%% BEGIN PROCEDURE
\section{Procedure}
A multidimentional array insertion sort can be implemented in a multitude of languages using the pseudocode provided in Algorithm 1.
\newline
\textbf{Insertion Sort Pre-Condition}: A is an unsorted non-empty array of non-empty arrays containing a comparable data type with a natural order such that v is an index value of the inner array.
\newline
\textbf{Insertion Sort Post-Condition}: A' is a permutation of A that is in strictly non-increasing order.
\begin{algorithm}
\caption {\textsc{Insertion-Sort}(A, v)}
\label{algo:insertionsort}
\begin{algorithmic}[1]
\Procedure{Insertion-Sort}{A, v}
\If{$A.length < 2$}
\State{\Return{$A$}}
\EndIf
\State{$i = 2$}
\While{$i$ upto $A.length$}
	\State{$key = A[i][v]$}
	\State{$a = A[i]$}
	\State{$j = i - 2$}
	\While{$j$ downto $1$ and $key > A[j][v]$}
		\State{$A[j+1] = A[j]$}
		\State{$j = j - 1$}
	\EndWhile
	\State{$A[j+1] = a$}
	\State{$i = i + 1$}
\EndWhile
\State{\Return{$A$}}
\EndProcedure
\end{algorithmic}
\end{algorithm}
\newline
\textbf{Insertion Sort Outer-Loop Invariant}: The subarray A'[1 ... i - 1] contains all the same elements as the subarray A[1 ... i - 1].
\newline 
\textbf{Insertion Sort Outer-Loop Initialization}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same one element.
\newline
\textbf{Insertion Sort Outer-Loop Maintenance}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same elements, although they may be in different orders.
\newline
\textbf{Insertion Sort Outer-Loop Termination}: When the outer-loop terminates, i = A.length, which implies that the entire array has been traversed and the guard has been negated. The negation of the guard implies that A'[1 ... i - 1] contains all the elements in A[1 ... i - 1].
\newline
\newline
\textbf{Insertion Sort Inner-Loop Invariant}: A'[1 ... j] is sorted in strictly non-decreasing order.
\newline
\textbf{Insertion Sort Inner-Loop Initialization}: Before the first iteration of the loop, j = 1, meaning the subarray A'[1 ... j] contains exactly one element, which is already sorted.
\newline
\textbf{Insertion Sort Inner-Loop Maintenance}: At the beginning of each iteration of the loop the inner-loop invariant holds because j counts down from i, and A'[j+1] is swapped with A'[j] only if A'[j+1] is less than A[j].
\newline
\textbf{Insertion Sort Inner-Loop Termination}: The negation of the guard implies that j = A.length and that A'[1 ... j] has been entirely traversed and sorted in strictly non-decreasing order, which maintains the inner-loop invariant. 
\newline
\newline
\textbf{Insertion Sort Conclusion}: The termination of both the inner and outer loops implies that the entire array has been traversed, A' is a permutation of A containing all the same elements in strictly non-decreasing order. This satisfies the post condition.
\newline
\newline

A greedy solution to a 0/1 knapsack problem can be implemented in a variety of languages using the pseudocode in Algorithm 2.
\newline
\textbf{Greedy Knapsack Pre-Condition}: $Weights$ and $Prices$ both have in them $n$ number of elements
\newline
\textbf{Greedy Knapsack Post-Condition}: The returned value will be a reasonable solution for the largest value of price combinations such that the aggregate of the corresponding weights does not exceed knapsack capacity $c$.
\begin{algorithm}
\caption {\textsc{Greedy-Knapsack}(n, Weights, Prices, c)}
\label{algo:greedy}
\begin{algorithmic}[1]
\Procedure{Greedy-Knapsack}{n, Weights, Prices, c}
\If{$n == 1$ and $Weights[1] <= c$}
	\State{\Return{$Prices[1]$}}
\EndIf
\If{$n <= 0$}
	\State{\Return{$0$}}
\EndIf
\State{$profit = 0$}
\State{$ratio = new Array$}
\For{$v = 1$ upto $n$}
	\State{$a = Prices[v] / Weights[v]$}
	\State{$ratio[v] = a$}
\EndFor
\State{$ratio = Insertion-Sort(ratio, 1)$}
\State{$i = 1$}
\While{$c > 0$ and $i < n$}
	\If{$c - ratio[i][3] >= 0$}
		\State{$profit = profit + ratio[i][2]$}
		\State{$c = c - ratio[i][3]$}
	\EndIf
	\State{$i = i + 1$}
\EndWhile
\State{\Return{$profit$}}
\EndProcedure
\end{algorithmic}
\end{algorithm}
\newline
\textbf{Greedy Solution First-Loop Invariant}: ratio[1 ... v-1] has the same number of non-null elements in it as both Prices[1 ... v-1] and Weights[1 ... v-1].
\newline
\textbf{Greedy Solution First-Loop Invariant Initialization}: The invariant holds true before the first iteration of the loop because $v=1$, and ratio[1 ... v-1], Prices[1 ... v-1] and Weights[0 ... v-1] all do not exist, and is therefore vacuously true.
\newline
\textbf{Greedy Solution First-Loop Invariant Maintenance}: At the beginning of each iteration of the loop the invariant holds true because each time the loop runs, v is incremented by 1, and exactly 1 element is added to ratio, therefore ratio[0 ... v-1] will have the exact same number of elements as both Prices[1 ... v-1] and Weights[1 ... v-1].
\newline
\textbf{Greedy Solution First-Loop Invariant Termination}: After the termination of the loop the invariant holds true because v has been incremented by 1 during each iteration of the loop, and ratio has gained 1 element during each iteration of the loop, therefore ratio[1 ... v-1] contains the exact same number of elements as both Prices[1 ... v-1] and Weights[1 ... v-1].
\newline
\newline
\textbf{Greedy Solution Second-Loop Invaraint}: The aggregate of the weights corresponding to the prices included in $profit$ are less than or equal to $c$.
\newline
\textbf{Greedy Solution Second-Loop Invariant Initialization}: Before the first iteration of the loop the invariant holds true vacuously because no prices have been included in $profit$.
\newline
\textbf{Greedy Solution Second-Loop Invariant Maintenance}: At the beginning of each iteration of the loop the invariant holds true because only items whose weight is less than or equal to the size of $c$ are added to $profit$.
\newline
\textbf{Greedy Solution Second-Loop Invariant Termination}: After the termination of the loop, the invariant holds true because only items whose weight was less than or equal to the value of $c$ have been added to $profit$.
\newline
\newline

@@

\textbf{Dynamic Solution Second Inner-Loop Invariant Termination}: After the termination of the loop the invariant holds true because the values in tab have never been decremeneted, only incremented.

@@

\subsection{Problems Encountered}
One major issue encountered during the development of the dynamic programming solution was that when the second pair of loops would make calls to any of ArrayLists (including $tab$, which at the time was also an ArrayList implementation) there would be an out of bounds error because of a negative index value. The out of bounds error was eminating from the conditional in the second nested loop. The cause for this error is still unknown. To aid the issue $tab$ was changed from a nested ArrayList to a primitive nested Java array.

@@

%% BEGIN EXPERIMENTAL ANALYSIS
\section{Experimental Analysis}
\begin{figure}[!]
\begin{center}
\includegraphics[scale=.55]{test-results.pdf}
\end{center}
\captionof{figure}{Java Implementation of Greedy and Dynamic Solutions to 0/1 Knapsack Run-Time Analysis}
\label{fig:runtime}
\end{figure}
The insertion sort, greedy solution and dynamic solution demonstrated in Algorithm 1, 2 and 3 was implemented in Java and executed on an HP SpectreXT TouchSmart with a 4 core Intel i7 Ivybridge processor clocked at 1.9GHz running Ubuntu Gnome 14.10 64-bit with 8GB of DDR3 RAM.
The expected growth of insertion sort as the number of elements (n) grows large can be represented as $ \theta{n^2} $ where $ \theta{} $ represents the asympotically tightly bound running time. \cite{textbook}

The greedy solution also follows a similair $ \theta{n^2} $ run time, see Figure 1. This is because an insertion sort was chosen to implement as the sorting algorithm used in the greedy solution. An insertion sort was chosen for the sorting algorithm because, while it is not the fatest, requires less auxillary space. While run time is an important factor, the space required to run the algorithm is also very important, and should not be neglacted. As seen in Table 2, the dynamic solution unavoidably requried too much auxillary space and Java gave an error when values of $n$ became incresingly large. It also important to note that the data graphed in Figure 1 was executed while the machine executed superfolous commands such as checking for system updates. To aid this issue, averages of 10 trails were taken and used in Figure 1.

The run time for the dynamic solution as seen in figure one $ \theta{n*c} $ where n is the number of elements and c is the size of the knapsack. In this implementation for values of $n$ greater than or equal to 100, a value of $2n$ was used for $c$. That would make the run time of the specific implementation of the dynamic solution $2n^2$ but after dropping constants becomes $ \theta{n^2} $. In Figure 1, the reason that the dynamic solution takes so much more time to complete is because there is a constant of 2 for $n^2$ which can be attributed to doubling the real time it took to return the results. Because the dynamic solution works by storing answers originally found in auxillary space, the values of $n$ cannot be very high without causing a Java error, as seen in Table 2.

In Table 3 there looks to be a negative correlation between the size of $n$ and the percent error. This negative correlation can be attributed to the fact that the percent error formula takes the difference between the two results and divides it by the correct result. For example, if the result for a greedy solution was 2 and the result for the dynamic solution was 3 the formula would read $ \frac{2}{3} * 100 $ rendering a 66.66\% error, however the difference between the two is only 2. If the values were 101 and 103 the pecent error formula would read $ \frac{2}{103} * 100 $ and renders 1.9\% error. The second value is substantially lower, even though the difference between the two numbers remains the same.

@@

%% BEGIN CONCLUSION
\section{Conclusion}
%%TODO: write this

@@

%% BEGIN REFERNCES
\section*{References}

%% \nocite{*}
%% \bibliographystyle{IEEEtran}
%% \bibliography{bib}

>>>>

%% BEGIN ABSTRACT
\begin{abstract}
The present study investigates two implementations for a 0/1 Knapsack problem, a greedy solution, and a dynamic solution. The dynamic solution will always result in the correct answer but takes $\theta{n*m}$ time to complete. The greedy solution will generally result in a reasonable answer, and even sometimes the correct answer, and take $\theta{n}$ time to complete. Which solution to use will depend on what is most important to the problem at hand, speed, or accuracy.

@@

%% BEGIN MOTIVATION
\section{Motivation}
A 0/1 Knapsack problem is most often described as how does a thief steal the most valuable items with the amount of space he has available to carry them. However, the applications of a 0/1 Knapsack far beyond just this. When solved, a 0/1 Knapsack problem will render the optimal solution, whether it be how one should construct their schedule so they can get the most done in one day, or how to pack a truck such that the fewest amount of trips are taken.

@@

%% BEGIN BACKGROUND
\section{Background}
Unfortunately, solving a 0/1 Knapsack problem often times proves difficult because in order to find the best solution, all possible solutions must be attempted, a method known as brute force. It is not practical to brute force a problem of this kind because it would take $ \theta{n!} $ to solve, which is unacceptable for large sets of numbers. However, there are other methods beside brute force that allow for the correct or 'good-enough' solution to be found. One of these methods is known as dynamic programming, which trades off speed for space. With dynamic programming the correct solution can be found in $ \theta{n*m} $ time, where n is the amount of items and m is the size of the knapsack. Another method for solving this type of problem is called a greedy algorithm, which works by making the best decision on which item to take locally. A greedy solution to the 0/1 Knapsack problem traditionally runs in $ \theta{n} $ time. Unfortunately a greedy solution is not always correct and therefore highly unreliable.

@@

%% BEGIN PROCEDURE
\section{Procedure}
A multidimensional array insertion sort can be implemented in a multitude of languages using the pseudocode provided in Algorithm 1.
\newline
\textbf{Insertion Sort Pre-Condition}: A is an unsorted non-empty array of non-empty arrays containing a comparable data type with a natural order such that v is an index value of the inner array.
\newline
\textbf{Insertion Sort Post-Condition}: A' is a permutation of A that is in strictly non-increasing order.
\begin{algorithm}
\caption {\textsc{Insertion-Sort}(A, v)}
\label{algo:insertionsort}
\begin{algorithmic}[1]
\Procedure{Insertion-Sort}{A, v}
\If{$A.length < 2$}
\State{\Return{$A$}}
\EndIf
\State{$i = 2$}
\While{$i$ upto $A.length$}
	\State{$key = A[i][v]$}
	\State{$a = A[i]$}
	\State{$j = i - 2$}
	\While{$j$ downto $1$ and $key > A[j][v]$}
		\State{$A[j+1] = A[j]$}
		\State{$j = j - 1$}
	\EndWhile
	\State{$A[j+1] = a$}
	\State{$i = i + 1$}
\EndWhile
\State{\Return{$A$}}
\EndProcedure
\end{algorithmic}
\end{algorithm}
\newline
\textbf{Insertion Sort Outer-Loop Invariant}: The subarray A'[1 ... i - 1] contains all the same elements as the subarray A[1 ... i - 1].
\newline 
\textbf{Insertion Sort Outer-Loop Initialization}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same one element.
\newline
\textbf{Insertion Sort Outer-Loop Maintenance}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same elements, although they may be in different orders.
\newline
\textbf{Insertion Sort Outer-Loop Termination}: When the outer-loop terminates, i = A.length, which implies that the entire array has been traversed and the guard has been negated. The negation of the guard implies that A'[1 ... i - 1] contains all the elements in A[1 ... i - 1].
\newline
\newline
\textbf{Insertion Sort Inner-Loop Invariant}: A'[1 ... j] is sorted in strictly non-decreasing order.
\newline
\textbf{Insertion Sort Inner-Loop Initialization}: Before the first iteration of the loop, j = 1, meaning the subarray A'[1 ... j] contains exactly one element, which is already sorted.
\newline
\textbf{Insertion Sort Inner-Loop Maintenance}: At the beginning of each iteration of the loop the inner-loop invariant holds because j counts down from i, and A'[j+1] is swapped with A'[j] only if A'[j+1] is less than A[j].
\newline
\textbf{Insertion Sort Inner-Loop Termination}: The negation of the guard implies that j = A.length and that A'[1 ... j] has been entirely traversed and sorted in strictly non-decreasing order, which maintains the inner-loop invariant. 
\newline
\newline
\textbf{Insertion Sort Conclusion}: The termination of both the inner and outer loops implies that the entire array has been traversed, A' is a permutation of A containing all the same elements in strictly non-decreasing order. This satisfies the post condition.
\newline
\newline

A greedy solution to a 0/1 knapsack problem can be implemented in a variety of languages using the pseudocode in Algorithm 2.
\newline
\textbf{Greedy Knapsack Pre-Condition}: $Weights$ and $Prices$ both have in them $n$ number of elements
\newline
\textbf{Greedy Knapsack Post-Condition}: The returned value will be a reasonable solution for the largest value of price combinations such that the aggregate of the corresponding weights does not exceed knapsack capacity $c$.
\begin{algorithm}
\caption {\textsc{Greedy-Knapsack}(n, Weights, Prices, c)}
\label{algo:greedy}
\begin{algorithmic}[1]
\Procedure{Greedy-Knapsack}{n, Weights, Prices, c}
\If{$n == 1$ and $Weights[1] <= c$}
	\State{\Return{$Prices[1]$}}
\EndIf
\If{$n <= 0$}
	\State{\Return{$0$}}
\EndIf
\State{$profit = 0$}
\State{$ratio = new Array$}
\For{$v = 1$ upto $n$}
	\State{$a = Prices[v] / Weights[v]$}
	\State{$ratio[v] = a$}
\EndFor
\State{$ratio = Insertion-Sort(ratio, 1)$}
\State{$i = 1$}
\While{$c > 0$ and $i < n$}
	\If{$c - ratio[i][3] >= 0$}
		\State{$profit = profit + ratio[i][2]$}
		\State{$c = c - ratio[i][3]$}
	\EndIf
	\State{$i = i + 1$}
\EndWhile
\State{\Return{$profit$}}
\EndProcedure
\end{algorithmic}
\end{algorithm}
\newline
\textbf{Greedy Solution First-Loop Invariant}: ratio[1 ... v-1] has the same number of non-null elements in it as both Prices[1 ... v-1] and Weights[1 ... v-1].
\newline
\textbf{Greedy Solution First-Loop Invariant Initialization}: The invariant holds true before the first iteration of the loop because $v=1$, and ratio[1 ... v-1], Prices[1 ... v-1] and Weights[0 ... v-1] all do not exist, and is therefore vacuously true.
\newline
\textbf{Greedy Solution First-Loop Invariant Maintenance}: At the beginning of each iteration of the loop the invariant holds true because each time the loop runs, v is incremented by 1, and exactly 1 element is added to ratio, therefore ratio[0 ... v-1] will have the exact same number of elements as both Prices[1 ... v-1] and Weights[1 ... v-1].
\newline
\textbf{Greedy Solution First-Loop Invariant Termination}: After the termination of the loop the invariant holds true because v has been incremented by 1 during each iteration of the loop, and ratio has gained 1 element during each iteration of the loop, therefore ratio[1 ... v-1] contains the exact same number of elements as both Prices[1 ... v-1] and Weights[1 ... v-1].
\newline
\newline
\textbf{Greedy Solution Second-Loop Invariant}: The aggregate of the weights corresponding to the prices included in $profit$ are less than or equal to $c$.
\newline
\textbf{Greedy Solution Second-Loop Invariant Initialization}: Before the first iteration of the loop the invariant holds true vacuously because no prices have been included in $profit$.
\newline
\textbf{Greedy Solution Second-Loop Invariant Maintenance}: At the beginning of each iteration of the loop the invariant holds true because only items whose weight is less than or equal to the size of $c$ are added to $profit$.
\newline
\textbf{Greedy Solution Second-Loop Invariant Termination}: After the termination of the loop, the invariant holds true because only items whose weight was less than or equal to the value of $c$ have been added to $profit$.
\newline
\newline

@@

\textbf{Dynamic Solution Second Inner-Loop Invariant Termination}: After the termination of the loop the invariant holds true because the values in tab have never been decremented, only incremented.

@@

\subsection{Problems Encountered}
One major issue encountered during the development of the dynamic programming solution was that when the second pair of loops would make calls to any of ArrayLists (including $tab$, which at the time was also an ArrayList implementation) there would be an out of bounds error because of a negative index value. The out of bounds error was emanating from the conditional in the second nested loop. The cause for this error is still unknown. To aid the issue $tab$ was changed from a nested ArrayList to a primitive nested Java array.

@@

%% BEGIN EXPERIMENTAL ANALYSIS
\section{Experimental Analysis}
\begin{figure}[!]
\begin{center}
\includegraphics[scale=.55]{test-results.pdf}
\end{center}
\captionof{figure}{Java Implementation of Greedy and Dynamic Solutions to 0/1 Knapsack Run-Time Analysis}
\label{fig:runtime}
\end{figure}
The insertion sort, greedy solution and dynamic solution demonstrated in Algorithm 1, 2 and 3 was implemented in Java and executed on an HP SpectreXT TouchSmart with a 4 core Intel i7 Ivybridge processor clocked at 1.9GHz running Ubuntu Gnome 14.10 64-bit with 8GB of DDR3 RAM.
The expected growth of insertion sort as the number of elements (n) grows large can be represented as $ \theta{n^2} $ where $ \theta{} $ represents the asymptotically tightly bound running time. \cite{textbook}

The greedy solution also follows a similar $ \theta{n^2} $ run time, see Figure 1. This is because an insertion sort was chosen to implement as the sorting algorithm used in the greedy solution. An insertion sort was chosen for the sorting algorithm because, while it is not the fastest, requires less axillary space. While run time is an important factor, the space required to run the algorithm is also very important, and should not be neglected. As seen in Table 2, the dynamic solution unavoidably required too much axillary space and Java gave an error when values of $n$ became increasingly large. It also important to note that the data graphed in Figure 1 was executed while the machine executed superfluous commands such as checking for system updates. To aid this issue, averages of 10 trails were taken and used in Figure 1.

The run time for the dynamic solution as seen in figure one $ \theta{n*c} $ where n is the number of elements and c is the size of the knapsack. In this implementation for values of $n$ greater than or equal to 100, a value of $2n$ was used for $c$. That would make the run time of the specific implementation of the dynamic solution $2n^2$ but after dropping constants becomes $ \theta{n^2} $. In Figure 1, the reason that the dynamic solution takes so much more time to complete is because there is a constant of 2 for $n^2$ which can be attributed to doubling the real time it took to return the results. Because the dynamic solution works by storing answers originally found in axillary space, the values of $n$ cannot be very high without causing a Java error, as seen in Table 2.

In Table 3 there looks to be a negative correlation between the size of $n$ and the percent error. This negative correlation can be attributed to the fact that the percent error formula takes the difference between the two results and divides it by the correct result. For example, if the result for a greedy solution was 2 and the result for the dynamic solution was 3 the formula would read $ \frac{2}{3} * 100 $ rendering a 66.66\% error, however the difference between the two is only 2. If the values were 101 and 103 the percent error formula would read $ \frac{2}{103} * 100 $ and renders 1.9\% error. The second value is substantially lower, even though the difference between the two numbers remains the same.

@@

%% BEGIN CONCLUSION
\section{Conclusion}
Although in this specific implementation, both the dynamic solution and the greedy solution had a run-time of $\theta{n^2}$, it can be possible for the dynamic solution to have a run-time of $\theta{n*c}$ and the greedy solution to have a run-time of $\theta{n}$. The best algorithm to use for a given problem is really a subjective decision and depends on two things, 1. how accurate does the answer to the problem really need to be, and 2. what is the time requirement to find out what the answer is.

@@

%% BEGIN REFERNCES
%%\section*{References}

\nocite{*}
\bibliographystyle{IEEEtran}
\bibliography{bib}
